\vspace{-1.5em} \bf Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-06-08 paper.tex
Partial update

>>>>

Genetic data for SARS-CoV-2 contains substantial amounts of missingness,
(nucleotides coded as `unknown' or masked out
as described in Section~\ref{sec:data_preprocessing})
and it is important to accounting for this missinging in a systematic way.
In \texttt{sc2ts} we automatically impute missing data as samples are added
into the ARG, using the LS model.
To do this, we assume that sites with missing nucleotides are uninformative
to the path probability, by setting the emission probability from any
state $(A,C,G,U,-)$ to `unknown' equal to 1.
Once the most likely copying path is determined, we then attach the
sample to the ARG (Figure~\ref{fig:overview_sc2ts} and following section).
For each sample attached to the ARG, we encode its nucleotide sequence by
recoding mutations where it differs from the nucleotide sequence
of its attachment node, but (importantly) ignoring sites with missing data.
Thus, we impute the
recording mutations

% As a result, we had to consider missingness in our implementation of
% A varying degree of missingness is present in the genetic data, that is, for a
% given sample, a subset of nucleotides may be coded as `unknown' at a set of
% positions. As a result, we had to consider missingness in our implementation of
% the LS Viterbi algorithm. One approach would be to `impute' all missing data
% beforehand using external software to hard-call or define a distribution over
% the possible missing nucleotides and use that in the determination of path
% probabilities.

% Another, which we carry out within \texttt{sc2ts}, is to assume
% sites with missing nucleotides are uninformative to the path probability, and
% impute them as each focal sequence is analysed, on the fly. We achieve this by
% setting the emission probability from any state $(A,C,G,U,-)$ to `unknown' as
% 1. Then, after determining a most likely copying path, we can impute these
% missing data by simply copying nucleotides present on the most likely copying
% path at the sites of missing data in the focal sequence. Note that by
% iteratively imputing missing data in focal sequences in this way, the reference
% panel being used for any newly observed focal sequence is guaranteed to be
% devoid of missingness.
